\bf Range Analysis for the IonMonkey JavaScript engine

### 2012-05-02 project-writings/paper.tex
Some fixes.

The core of our analysis system is essentially an implementation of
Gough and Klaeren's static range analysis algorithm for bounds check
elimination.
We identify our technical contributions as:
\begin{itemize}
\item An implementation of Gough and Klaeren's range analysis in a
  modern industrial-strength SSA-based compiler.
\item A different algorithm for inserting and removing $\beta$ nodes
  that used by Gough and Klaeren that takes advantage of the data
  structures of a modern SSA compiler.
\item To facilitate the above, extending IonMonkey with a fast
  (constant time, one integer per basic block memory overhead) 
  mechanism for determining whether a block is dominated by another.
\item Extending the analysis to produce the information needed for
  overflow check elimination by differentiating between
  ``maximum/minimum 32-bit int values'' and ``infinity'' (possible
  overflow).
\item The observation that when doing range computation with the above
  distinction, the input ranges can be always be ``squashed'' to only
  finite ranges, since if overflow had occurred, the code would have
  bailed out.
\item An implementation of overflow check elimination using the above
  analyses.
\end{itemize}

@@

An important source of range information that requires care to take
advantage of is conditional branches. Consider the code in Figure
\ref{fig:branch}. All three arithmetic operations in this code can be
proven to not overflow, but to do this it is not sufficient to simply
associate each name with a range, since the information differs
between basic blocks. An obvious approach is to instead associate
ranges with name/basic block pairs, allowing representation of
flow-sensitive range information. This solution, however, is
unsatisfying. One of the major benefits of SSA-style analysis is
eliminating the need for this sort of tracking; by giving each
computation a unique name, SSA reduces the need for tracking
information about variables across time; adopting this approach would
undo this.

@@

\begin{lstlisting}[language=Javascript,
                   caption={Code that requires information about branches to properly analyze},
                   label={fig:branch}, float=ht]
var y;
if (x < 0) {
    y = x + 2000000000;
} else {
    if (x < 1000000000) {
        y = x*2;
    } else {
        y = x - 3000000000;
    }
}
\end{lstlisting}

\begin{lstlisting}[language=Javascript,
                   caption={Code from Figure \ref{fig:branch}, in XSA form},
                   label={fig:xsa}, float=ht]
if (x < 0) {
    x1 = Beta(x, [INT_MIN, -1]);
    y1 = x1 + 2000000000;
} else {
    x2 = Beta(x, [0, INT_MAX]);
    if (x2 < 1000000000) {
        x3 = Beta(x2, [INT_MIN, 999999999]);
        y2 = x3*2;
    } else {
        x4 = Beta(x2, [1000000000, INT_MAX]);
        y3 = x4 - 3000000000;
    }
    y4 = Phi(y2, y3);
}
y = Phi(y1, y4);
\end{lstlisting}

@@

After inserting a $\beta$ node, we then want to propagate this range information to all uses
of the variable where the branch is guaranteed to have been taken;
this is accomplished by replacing all uses of original instruction
that are dominated by the new $\beta$ instruction with the $\beta$. To
do this, we take advantage of IonMonkey's use chains to efficiently
iterate over all uses of an instruction.

@@

Recomputation of the possible range of an instruction is fairly simple
and proceeds by the rules given in Figure \ref{fig:range_anal}. The
key insight we had in these equations is that while the ranges of the
inputs to an operation may be infinite (that is, they may have
overflowed), this possibility can be ignored when using the ranges as
inputs. This is because had the computation overflowed, it would have
been detected and IonMonkey would bail out to non-integer code. We
know that if code that assumes the input is an integer is still being
run, then the operation must not have overflowed, and thus must have a
finite range. What this means for the range analysis equations is that
we can always ``squash'' the ranges of inputs by replacing positive
and negative infinite with the maximum and minimum 32-bit integers,
respectively.

@@

\subsection{Array Bounds Check Elimination}
We do not perform general array bounds check elimination. To do this
fully would require symbolic range analysis. Since IonMonkey already
does hoisting of bounds checks out of loops, we determined that the
effort would probably not be worth the gain.

>>>>

The core of our analysis system is essentially an implementation of
Gough and Klaeren's static range analysis algorithm for bounds check
elimination.
We identify our technical contributions as:
\begin{itemize}
\item An implementation of Gough and Klaeren's range analysis in a
  modern industrial-strength SSA-based compiler.
\item A different algorithm for inserting and removing $\beta$ nodes
  than  used by Gough and Klaeren that takes advantage of the data
  structures of a modern SSA compiler.
\item To facilitate the above, extending IonMonkey with a fast
  (constant time, one integer per basic block memory overhead) 
  mechanism for determining whether a block is dominated by another.
\item Extending the analysis to produce the information needed for
  overflow check elimination by differentiating between
  ``maximum/minimum 32-bit int values'' and ``infinity'' (possible
  overflow).
\item The observation that when doing range computation with the above
  distinction, the input ranges can be always be ``squashed'' to only
  finite ranges, since if overflow had occurred, the code would have
  bailed out.
\item An implementation of overflow check elimination using the above
  analyses.
\end{itemize}

@@

An important source of range information that requires care to take
advantage of is conditional branches. Consider the code in Listing
\ref{fig:branch}. All three arithmetic operations in this code can be
proven to not overflow, but to do this it is not sufficient to simply
associate each name with a range, since the information differs
between basic blocks. An obvious approach is to instead associate
ranges with name/basic block pairs, allowing representation of
flow-sensitive range information. This solution, however, is
unsatisfying. One of the major benefits of SSA-style analysis is
eliminating the need for this sort of tracking; by giving each
computation a unique name, SSA reduces the need for tracking
information about variables across time; adopting this approach would
undo this.

@@

\begin{lstlisting}[language=Javascript,
                   caption={Code that requires information about branches to properly analyze},
                   label={fig:branch}, float=ht]
if (x < 0) {
    y = x + 2000000000;
} else {
    if (x < 1000000000) {
        y = x*2;
    } else {
        y = x - 3000000000;
    }
}
\end{lstlisting}

\begin{lstlisting}[language=Javascript,
                   caption={Code from Listing \ref{fig:branch}, in XSA form},
                   label={fig:xsa}, float=ht]
if (x < 0) {
    x1 = Beta(x, [INT_MIN, -1]);
    y1 = x1 + 2000000000;
} else {
    x2 = Beta(x, [0, INT_MAX]);
    if (x2 < 1000000000) {
        x3 = Beta(x2, [INT_MIN, 999999999]);
        y2 = x3*2;
    } else {
        x4 = Beta(x2, [1000000000, INT_MAX]);
        y3 = x4 - 3000000000;
    }
    y4 = Phi(y2, y3);
}
y = Phi(y1, y4);
\end{lstlisting}

@@

After inserting a $\beta$ node, we then want to propagate this range information to all uses
of the variable where the branch is guaranteed to have been taken;
this is accomplished by replacing all uses of original instruction
that are dominated by the new $\beta$ instruction with the $\beta$. To
do this, we take advantage of IonMonkey's use chains to efficiently
iterate over all uses of an instruction and modify them in place.

@@

Recomputation of the possible range of an instruction is fairly simple
and proceeds by the rules given in Figure \ref{fig:range_anal} (many
operations that we support or omitted from the figure). The key
insight we had in these equations is that while the ranges of the
inputs to an operation may be infinite (that is, they may have
overflowed), this possibility can be ignored when using the ranges as
inputs. This is because had the computation overflowed, it would have
been detected and IonMonkey would bail out to non-integer code. We
know that if code that assumes the input is an integer is still being
run, then the operation must not have overflowed, and thus must have a
finite range. What this means for the range analysis equations is that
we can always ``squash'' the ranges of inputs by replacing positive
and negative infinite with the maximum and minimum 32-bit integers,
respectively.

@@

\subsection{Array Bounds Check Elimination}
We do not perform general array bounds check elimination. To do this
would in many cases require symbolic range analysis. Since IonMonkey
already does hoisting of bounds checks out of loops, we determined
that the effort would probably not be worth the gain.
